The evolution of the transcriptome of mammalian placentas

### 2016-09-09 manuscript/manuscript.Rnw
gut the manuscript for rewriting

\begin{abstract}
  \textbf{Background} The placenta is arguably the most
  anatomically variable organ in mammals.

  \textbf{Results} We measured the expression profiles of
  \Sexpr{star.logs[,length(unique(file))]} term placentas of
  \Sexpr{star.logs[,length(unique(species))]} species of mammals. We
  found lineage-specific differential expression of prolactin,
  relaxin, and galectins as well as increased expression of the
  pre-eclampsia associated genes \textit{ADAM12}, \textit{CRH}, and
  \textit{PAPPA} in humans. We also identified genes which are
  expressed in all placentas, including genes with immune-modulating
  properties (\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and
  \textit{S100A10}) and cell-cell interactions (\textit{LAMC1},
  \textit{LUM}, and \textit{LGALS1}).

  \textbf{Conclusions} A diverse array of genes are differentially
  expressed and are likely to play a role in the anatomical
  variability of the placenta in the placentas of mammalian lineages.
  Furthermore, multiple genes which are associated with pre-eclampsia
  (\textit{CRH}, \textit{ADAM12}, \textit{KISS1}, \textit{PAPPA},
  \textit{IL1RL1}, and \textit{PLAC1}) show increased expression on
  the human lineage. Still, using the strictest filtering strategy
  we find a core set of only
  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
  genes is essential for the proper functioning of the placenta in all
  species analyzed.

@@

\section{Background}

The main role of the placenta in viviparous mammals is the transfer of
nutrients, oxygen, and waste between mother and offspring. Although
these functions are conserved among mammals, the placenta is perhaps
the most variable organ in terms of structure and form
\cite{Mossman.1937:comparativemorphogenesisofthe,Moffett.ea2006:Immunologyofplacentationin}. Species differ in
the shape of attachment (i.e., discoid, diffuse, cotyledonary,
zonary), degree of intimacy between fetal tissues and maternal tissues
(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
interdigitation of maternal-fetal tissue (i.e., villous, folded,
trabecular, lamellar, labrynthine)
\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
mammals differ drastically in the number and size of their offspring,
as well as the length of gestation, traits often associated with
specific aspects of placental morphology
\cite{Capellini.ea2011:Placentationandmaternalinvestment}
(\cref{tab:gestational_periods}).

The placenta arose at some point before the origin of extant mammals,
a group that includes marsupials, monotremes, and eutherian or
“placental mammals”. The term “placental mammals” is misleading, as
extant monotremes (such as \textit{Ornithorhynchus anatinus}, the
platypus) possess a rudimentary placenta, as do marsupials, although
the direct apposition to maternal tissues lasts for a proportionally
shorter time during gestation
\cite{Selwood.ea2006:Trophoblastandhypoblastin,Renfree2010:ReviewMarsupialsplacentalmammals}.
Marsupials also differ from eutherian mammals in that most possess
only a choriovitelline placenta, produced from the fusion of the
chorion and yolk sac, during most of gestation. Although some
eutherian mammals may possess a choriovitelline placenta at some point
in development, the major site of nutrient exchange in placentals
occurs via the chorioallantoic placenta, produced from the fusion of
the chorion and allantois
\cite{Mossman.1987:vertebratefetalmembranes}. It is unclear whether
the first placenta was more like that of marsupials or eutherians;
however, using phylogenetic methods, the ancestral eutherian was
reconstructed as possessing a relatively invasive hemochorial
placenta, as in humans and mouse
\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.

Although the morphology of the mammalian placenta has been well
characterized in a number of species
\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
Studies of certain model species (i.e., \textit{Homo sapiens},
\textit{Mus musculus}, \textit{Bos taurus}) have revealed a stunning
divergence in the genetics of the placenta, especially in later stages
of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
These include independent expansions of the prolactins in \mouse,
\textit{Spalax carmeli}, \textit{Spalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
in \cow, genes related to the thymus-specific serine protease
\textit{PRSS16} in \textit{Loxodona africana} (in this work),
 and the expansion of the growth hormone
and chorionic gonadotropin beta genes in anthropoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of \textit{SIGLEC6} to placental expression in \human
differs from both \textit{Pan troglodytes} and \mouse
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

One explanation for the divergent paths of the placental interface is
the distinct evolutionary interplay between maternal and fetal
strategies during pregnancy \cite{Haig1993:Geneticconflictsinhuman}.
The maternal strategy is to requisition an adequate degree of
nutrients to the fetus without risking the health of the mother, while
the fetal strategy via paternal genes is to extract as many resources
from the mother as possible. Therefore, an interplay between these
extremes can set up an intricate “arms race”, involving the imprinting
of parental alleles, expansion of gene families, and enlistment of
existing genes to expression at the maternal-fetal interface
\cite{Haig1993:Geneticconflictsinhuman}.

One symptom of this maternal-fetal conflict is its potential affect on
health
\cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. In
addition to having an invasive hemochorial placenta, \human possess a
deeply invasive trophoblast layer that can penetrate the myometrium of
the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
Defects in placental growth are known to be associated with the
devastating condition of pre-eclampsia, which can cause hypertension,
proteinuria, and has the potential to lead to death of the mother
and/or infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
Although cases have been documented in \textit{P. troglodytes}
\cite{Stout.ea1969:Glomerularcapillaryendothelialswelling}, and
\textit{Gorilla gorilla}
\cite{Thornton.ea1992:Convulsionsinpregnancyin}, pre-eclampsia has a
far higher incidence in \human pregnancies with as much as 4\% of
pregnancies being affected in the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. \textit{P.
  troglodytes} and \textit{G. gorilla} also have deeply invasive
trophoblasts, although the progression of invasion in great apes may
be slower than \human
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. However,
large-scale expression studies of great ape placentas have not been
conducted. The increasing affordability of next-generation sequencing
technology has enabled the large-scale study of specific tissues from
a wide array of species to address broader phylogenetic questions in
the evolution of gene expression
\cite{Brawand.ea2011:evolutionofgeneexpression}.

Here we use RNA-Seq
technology to sequence and quantify the mRNA environment of mammalian
placentas at term, including that of the \human, \textit{Pan paniscus}
(the sister species to \textit{P. troglodytes}), \textit{Ateles
  fusciceps}, \mouse, \galili, \carmeli, \textit{Canis familiaris},
\cow, \elephant, \textit{Dasypus novemcinctus}, and
\textit{Monodelphis domesticus}. These species represent four major
branches of eutherian mammals (Euarchontoglires, Laurasiatheria,
Afrotheria, and Xenartha) as well as marsupials. These placentas also represent a
wide array of placental morphologies as well as traits such as number
and size of offspring and length of gestation
(\cref{tab:gestational_periods}). We use this data, along with
existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual}, \textit{Equus caballus}
\cite{Coleman.ea2010:Structuralannotationofequine}, \textit{Ovis aries}
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \textit{Sus scrofa}
\cite{Ahanda.ea2012:Predictionofaltered3-} to address the major
differences between expression in marsupial and eutherian mammals at
term, identify the evolution of expression in important placental
genes such as hormones and galectins, identify novel lineage-specific
expression patterns, and quantify \human specific differences in gene
expression that may be related to obstetrical disorders such as
pre-eclampsia.

@@

\subsection{Exploration of Individual Transcriptome Data}

To determine enrichment levels of functional categories of genes in
highly-expressed transcripts in each transcriptome, we performed
functional annotation cluster analyses per transcriptome using topGO
\cite{topGO:2.20.0}. Gene Ontology clusters were considered
significantly enriched if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure \cite{Benjamini1995:bh_procedure} on from the
$p$ values obtained from the Fisher Exact Test were less than or equal
to 0.05. Enrichment tests were considered with two gene lists per
transcriptome: annotated genes with expression levels of the top 5\%
of the frequency distribution of \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_spalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris});
the top 100 genes with the highest \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_spalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
Due to the paucity of annotated gene ontology terms for some species,
gene lists were compared using the \human Gene Ontology using the 1:1
\human ortholog. Enrichment was considered out of the set of all 1:1
orthologs present in each species.

@@

\subsection{Primate Placenta Transcriptomes}

\begin{table}[htb]
\footnotesize
<<primate.genes,echo=FALSE,results="asis">>=
output.mentioned.genes.table(subset.table=quote(grepl("sapiens|fusciceps|paniscus",Species)),
                             human.symbol=TRUE
                             )
@ 
\caption{Notable genes ordered by species and \ac{FPKM}. Gene
  symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
  name corresponding to the human symbol. A complete table of 
  genes with \ac{FPKM} $\ge 50$ is located in the supplemental material }
\label{tab:notable_primate_genes}
\end{table}

\begin{table}[htb]
  \footnotesize
<<primate.lineage.specific,echo=FALSE,results="asis">>=
pri.all.lin.spec <- placenta.lineage.specific.table("pri.all")
setkey(pri.all.lin.spec,"Human Symbol")
@ 
\caption{\lineagetableexplanation{primates}{all other analyzed species}}
\label{tab:primate_lineage_specific}
\end{table}

We examined the expression of three primates, \human, \bonobo, and
\smonkey. The genes whose expressions are different on the primate
lineage in comparison to all other analyzed species are listed in
\cref{tab:primate_lineage_specific}. \textit{FSTL1} has significantly
higher expression in the primate lineage
(Supplemental \cref{fig:fstl1_evolution_tree},
$\text{FDR}=\Sexpr{pri.all.lin.spec["FSTL1",`FDR`]}$), is implicated
in BMP signaling, and has previously been implicated in inflammation
and invasiveness
\cite{Mouillet.ea2015:expressionandpost-transcriptionalregulation}. Of
non-primates, \horse shows the highest expression
%
\Sexpr{gene.expression(species.name="equus caballus",gene="FSTL1",complete=TRUE)}, 
%
which is still lower than the lowest primate, \smonkey
%
(\Sexpr{gene.expression(species.name="ateles fusciceps",gene="FSTL1",complete=TRUE)}). 
%
No other genes were significantly differentially expressed on this
lineage by t-test.

\subsubsection{\human Placenta Transcriptome}
\label{sec:human_placenta}

% \begin{table}
%   \footnotesize
% <<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
% output.top.genes.table(top.genes.table(species.name="homo sapiens"))
% @ 
% \caption{The top 50 genes by \ac{FPKM} in \human which are not
%   annotated as Mt\_rRNA, Mt\_tRNA, rRNA or a housekeeping gene. Gene
%   symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
%   name corresponding to the human symbol. A complete list of all genes
%   is in \cref{tab:all_homo_sapiens_placenta}.}
% \label{tab:top50_human_placenta}
% \end{table}

<<cgb.min.expression,results="hide",echo=FALSE,warnings=FALSE>>=
cgb.min.expression <- 
    min(sapply(paste0("CGB",c("","1","2","5","7","8")),
               function(x){gene.expression(species.name="homo sapiens",gene=x,units=FALSE)}))
cgb.sum.expression <- 
    sum(sapply(paste0("CGB",c("","1","2","5","7","8")),
               function(x){gene.expression(species.name="homo sapiens",gene=x,units=FALSE)}))
@

The gene with by far the highest \ac{FPKM} value
%
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
%
in \human is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within catarrhines (Supplemental \cref{fig:galectins_evolution_tree}). Another member of the growth
hormone gene family, \ac{GH2} (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the \human placenta
(Supplemental \cref{tab:all_homo_sapiens_placenta}).
% 
Eleven \acp{PSG} are highly-expressed
(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
greatest degree of expression %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
and the pseudogene \textit{PSG10P} showing the least %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~A) . %
These
genes are members of a hominoid-specific expansion of the \ac{CEA} family
and known to be expressed by the syncytiotrophoblast
\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
%
Genes involved in the production of estrogens (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
% 
) and progesterone (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
%
show exceptionally high expression in the human placenta, as well as %
%
\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
%
which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (Supplemental \cref{tab:all_homo_sapiens_placenta}). The genes
producing the beta subunits of the chorionic gonadotropins
(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
\textit{CGB7}, \textit{CGB8}) have reasonable expression
($\text{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
expression is on the order of the expression of
\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
%
%
Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. \textit{HBA1}, \textit{HBA2}, \textit{HBB},
and \textit{HBG2}), the maternally imprinted long noncoding RNA gene
\textit{H19}, genes coding for proteins required for growth and/or
cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
glucose transport (\textit{SLC2A1}) among others
(Supplemental \cref{tab:all_homo_sapiens_placenta}, Supplemental
\cref{tab:all_homo_sapiens_placenta}).

% \paragraph{Pre-eclampsia \human Lineage Specific Genes}

@@

\subsubsection{\bonobo Placenta Transcriptome}

% \begin{table}
%   \footnotesize
% <<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
% top.bonobo.genes <- top.genes.table(species.name="pan paniscus")
% output.top.genes.table(top.bonobo.genes)
% setkey(top.bonobo.genes,"Gene symbol")
% @ 
% \caption{\tableexplanation{\bonobo}{tab:all_pan_paniscus_placenta}}
% \label{tab:top50_bonobo_placenta}
% \end{table}

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \bonobo are shown in Supplemental \cref{tab:all_pan_paniscus_placenta}
with a list of all genes in Supplemental
\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
placenta differs somewhat from that of \human, as evidenced by the
makeup of the top 50 genes (Supplemental \cref{tab:all_pan_paniscus_placenta}). The highest expression
 in \bonobo is %
\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
%
which is involved in insulin growth factor signaling and angiogenesis
during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
\textit{PRG2} is also expressed at relatively high levels (though
comparatively much lower) in the human placenta
(\Sexpr{gene.expression("homo sapiens","PRG2")};
Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
related to the chorionic somatomammotropins
are also expressed at relatively high levels (\textit{GH1} 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
\textit{GH1} 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
\textit{GH2}
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
and \textit{GH2}
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
%
\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
Placental expression of the genes involved in estrogen and
progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
respectively) show lower expression in \bonobo as compared to \human %
(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
\cref{fig:estrogen_evolution_tree}; %
\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
\cref{fig:progesterone_evolution_tree} and
Supplemental \cref{tab:all_homo_sapiens_placenta}) although the expression of
\textit{HSD3B1} is fairly similar to that of \smonkey %
(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
In addition, \textit{CGA} is expressed in the bonobo placenta %
(\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
lower levels than \human as well (Supplemental \cref{tab:all_homo_sapiens_placenta}).
%
Two Ensembl-annotated beta subunits of the chorionic gonadotropin
genes are expressed in \bonobo, one related to \textit{CGB} and
another related to \textit{CGB7}; both are not highly expressed
(\cref{fig:cga_cgb_evolution_tree}). Among the other top 50 expressed
genes, there is high expression of keratins (e.g., \textit{KRT19},
\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2},
Supplemental \cref{fig:serpine_evolution_tree}), insulin growth factor
2 (\textit{IGF2}) and one of its binding proteins \textit{IGFBP3}, and
the hormone adrenomedullin (\textit{ADM}).

\subsubsection{\smonkey Transcriptome}

% \begin{table}
%   \footnotesize
% <<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
% top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
% output.top.genes.table(top.spider.monkey.genes)
% setkey(top.spider.monkey.genes,"Gene symbol")
% @ 
% \caption{\tableexplanation{\smonkey}{tab:all_ateles_fusciceps_placenta}}
% \label{tab:top50_spider_monkey_placenta}
% \end{table}

For the term placenta of \smonkey, assembled contigs from RNA-seq were
aligned using against \human and the best alignment was chosen as the
annotation for each assembled contig. In the term placenta of
\smonkey, the top 50 non-housekeeping genes
(Supplemental \cref{tab:all_spider_monkey_placenta}) include a possible regulator
of androgen receptor signaling
\cite{Sampson.ea2012:PAGE4positivityisassociated}, \textit{PAGE4} %
(\Sexpr{gene.expression("ateles_fusciceps","PAGE4")}), %
the pregnancy specific glycoprotein \textit{PSG8}
(\Sexpr{gene.expression("ateles_fusciceps","PSG8")}), %
only one member of the chorionic somatotropins, \textit{GH1} %
(\Sexpr{gene.expression("ateles_fusciceps","GH1")}), %
a few HLA genes (\textit{HLA-A}, \textit{HLA-F},
Supplemental \cref{tab:all_spider_monkey_placenta}), and multiple components of
annexin complexes
(\Sexpr{gene.expression("ateles_fusciceps","ANXA2",complete=TRUE)}, %
\Sexpr{gene.expression("ateles_fusciceps","S100P",complete=TRUE)}, %
\Sexpr{gene.expression("ateles_fusciceps","S100A11",complete=TRUE)}) %
which may play a role in angiogenesis, and certainly regulate
cell-cell interaction in the developing placenta
\cite{Liu.ea2015:AnnexinA2complexeswith}.
\Sexpr{gene.expression("ateles_fusciceps","AL365202.1",complete=TRUE)}
is also highly expressed, has no known function, and is duplicated
multiple times in \textit{Macaca mulatta} (Ensembl gene tree
\href{http://useast.ensembl.org/Multi/GeneTree/Image?gt=ENSGT00580000081348}{ENSGT00580000081348}).
All genes with expression greater than 50 \ac{FPKM} are listed in
Supplementary \cref{tab:all_ateles_fusciceps_placenta}.

\subsection{Rodentia Placenta Transcriptomes}

\begin{table}
  \footnotesize
<<rodentia.genes,echo=FALSE,results="asis">>=
output.mentioned.genes.table(subset.table=quote(grepl("musculus|galili|carmeli",Species)),
                             human.symbol=TRUE
                             )
@ 
\caption{Notable genes ordered by species and \ac{FPKM}. Gene
  symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
  name corresponding to the human symbol. A complete table of 
  genes with \ac{FPKM} $\ge 50$ is located in the supplemental material }
\label{tab:notable_non_primate_genes}
\end{table}

We examined the expression of three rodents, \mouse, \carmeli, and
\galili{}. We did not detect any significant changes on the Rodentia
lineage in comparison to all other analyzed species (Supplemental
\cref{tab:rodentia_lineage_specific}), likely because of a lack power
to overcome the multiple testing correction. Additional species and/or
samples in each of these lineages will be necessary to detect lineage
specific shifts.

\subsubsection{\mouse Placenta Transcriptome}

% \begin{table}[h]
%   \footnotesize
% <<top.50.mouse.expressed.genes,echo=FALSE,results="asis">>=
% output.top.genes.table(top.genes.table(species.name="mus musculus"))
% @ 
% \caption{\tableexplanation{\mouse}{tab:all_mus_musculus_placenta}}
% \label{tab:top50_mouse_placenta}
% \end{table}

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \mouse are shown in
Supplemental \cref{tab:all_mus_musculus_placenta} with a list of all genes with
$\text{FPKM} \ge 50$ in Supplemental
\cref{tab:all_mus_musculus_placenta}. The top expressed genes include
\Sexpr{gene.expression("mus_musculus","Mir6236",complete=TRUE)}, whose
target \textit{in silico} is the orphan G-coupled receptor
\textit{Mrgprg}
\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget},
\Sexpr{gene.expression("mus_musculus","H19",complete=TRUE)} and
\Sexpr{gene.expression("mus_musculus","Gm27786",complete=TRUE)}, both
of which are maternally imprinted (the later of which is a homolog of
human \textit{H19\_2}),
\Sexpr{gene.expression("mus_musculus","Igf2",complete=TRUE)}, numerous
cathepsins and a rodentia specific expansion of prolactin related
genes. There are also two miRNA which overlap with the of the 7S
signal recognition particle
(\Sexpr{gene.expression("mus_musculus","AC157822.1",complete=TRUE)}
and
\Sexpr{gene.expression("mus_musculus","AC099934.1",complete=TRUE)})
and numerous pseudogenes
(\Sexpr{gene.expression("mus_musculus","Gm11966",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm10736",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm10180",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm29216",complete=TRUE)}) which
appear to be highly expressed, however, the short read length (34--44)
in many samples is likely preventing unambiguous assignment of all
reads to the more lengthy complete transcript, so these shorter
transcripts should be treated with suspicion, and we will ignore them
in further discussion. The prolactin-related genes
(\Sexpr{gene.expression("mus_musculus","Prl3b1",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Prl2b1",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Prl8a9",complete=TRUE)}, and
\Sexpr{gene.expression("mus_musculus","Prl8a6",complete=TRUE)}) are
very highly expressed in \mouse in, with %
\Sexpr{combined.fpkm[species=="mus musculus" &
  grepl("Prl(\\d.\\d)?",gene_short_name), sum(mean_fpkm >= 1)]}
prolactin related genes having $\text{FPKM} \ge 1$
(\cref{fig:prolactin_evolution_tree}, Supplemental
\cref{fig:prolactin_receptor_evolution_tree}). The cathespins have
also expanded in rodents, with six showing high expression
(\Sexpr{gene.expression("mus_musculus","Ctsj",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Cts6",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Ctsr",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Ctsq",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Cts3",complete=TRUE)}, and
\Sexpr{gene.expression("mus_musculus","Ctsm",complete=TRUE)},
Supplemental \cref{tab:all_mus_musculus_placenta,fig:pag2_prss16_ctsl_psg_evolution_tree}).
Finally, \textit{Peg10}, an imprinted gene required for proper
formation of the placenta in mice is also highly expressed
(\Sexpr{gene.expression("mus_musculus","Peg10")}), as previously
reported \cite{Ono.ea2006:DeletionofPeg10imprinted}.

% \paragraph{\mouse Lineage Specific Genes}

\begin{table}
  \footnotesize
<<mouse.lineage.specific,echo=FALSE,results="asis">>=
mou.all.lin.spec <-
    placenta.lineage.specific.table("mou.all",
                                    lineage.table=human.mouse.lineage.specific.analysis,
                                    members=40,
                                    orderbyfc=TRUE,
                                    fdr.cutoff=0.001
                                    )
setkey(mou.all.lin.spec,"Human Symbol")
@ 
\caption{\lineagetableexplanationtypeiii{\mouse}{all other analyzed species}}
\label{tab:mouse_lineage_specific}
\end{table}

Using a mixed-effects linear model to account for repeated testing of
\human and \mouse, we identified multiple genes which are
differentially expressed in the \mouse lineage in comparison to all
other lineages (\cref{tab:mouse_lineage_specific}). \mouse
\textit{Gkn2} (FDR=\Sexpr{mou.all.lin.spec["GKN2",`FDR`]},
FC=\Sexpr{mou.all.lin.spec["GKN2",`Fold Change`]}) has been shown to
have a pattern of expression distinct from the \human pattern, and has
been suggested to be important for placental development
\cite{Fahlbusch.ea2015:Trophoblastexpressiondynamicsof}. Notable
mouse-specific increases in expression include include genes like
\textit{Slc38a4} (FDR=\Sexpr{mou.all.lin.spec["SLC38A4",`FDR`]},
FC=\Sexpr{mou.all.lin.spec["SLC38A4",`Fold Change`]}) and \textit{Afp}
(FDR=\Sexpr{mou.all.lin.spec["AFP",`FDR`]},
FC=\Sexpr{mou.all.lin.spec["AFP",`Fold Change`]}) which are implicated
in fetal growth restriction
\cite{Sharma.ea2016:Tpbpa-Cre-mediateddeletionofTFAP2C} and placenta
abnormalities in humans
\cite{Waller.ea1996:associationbetweenmaternalserum}, and
\textit{Cnx2} (FDR=\Sexpr{mou.all.lin.spec["CNX2",`FDR`]},
FC=\Sexpr{mou.all.lin.spec["CNX2",`Fold Change`]}) which is critical
for the exchange of metabolites in the placenta labyrinth
\cite{Gabriel.ea1998:Transplacentaluptakeofglucose}.

\subsubsection{\galili Placenta Transcriptome}

% \begin{table}[h]
%   \footnotesize
% <<top.50.n.galili.expressed.genes,echo=FALSE,results="asis">>=
% output.top.genes.table(top.genes.table(species.name="spalax galili"))
% @ 
% \caption{\tableexplanation{\galili}{tab:all_spalax_galili_placenta}}
% \label{tab:top50_n_galili_placenta}
% \end{table}

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \galili are shown in
Supplemental \cref{tab:all_n_galili_placenta} with a list of all highly expressed
genes in Supplemental \cref{tab:all_spalax_galili_placenta}.
Components of annexin complexes are highly expressed
(\Sexpr{gene.expression("spalax_galili","S100a6",complete=TRUE)},
\Sexpr{gene.expression("spalax_galili","S100a11",complete=TRUE)},
\Sexpr{gene.expression("spalax_galili","S100a10",complete=TRUE)},
and
\Sexpr{gene.expression("spalax_galili","Anxa2",complete=TRUE)}),
as are members of the thymosin family
(\Sexpr{gene.expression("spalax_galili","Tmsb4x",complete=TRUE)}
and
\Sexpr{gene.expression("spalax_galili","Tmsb10",complete=TRUE)}),
various cytoskeleton related proteins
(\Sexpr{gene.expression("spalax_galili","Actc1",complete=TRUE)},
\Sexpr{gene.expression("spalax_galili","Actg2",complete=TRUE)},
\Sexpr{gene.expression("spalax_galili","Acta2",complete=TRUE)},
\Sexpr{gene.expression("spalax_galili","Myl9",complete=TRUE)},
\Sexpr{gene.expression("spalax_galili","Vim",complete=TRUE)}), and
transmembrane protein of currently unknown function, 
\Sexpr{gene.expression("spalax_galili","Gm10134",complete=TRUE)}.

\subsubsection{\carmeli Placenta Transcriptome}

% \begin{table}[h]
%   \footnotesize
% <<top.50.s.carmeli.expressed.genes,echo=FALSE,results="asis">>=
% output.top.genes.table(top.genes.table(species.name="spalax carmeli"))
% @ 
% \caption{\tableexplanation{\carmeli}{tab:all_spalax_carmeli_placenta}}
% \label{tab:top50_s_carmeli_placenta}
% \end{table}

>>>>

\begin{abstract}
  \textbf{Introduction} The placenta is arguably the most
  anatomically variable organ in mammals.

  \textbf{Methods}

  \textbf{Results} We measured the expression profiles of
  \Sexpr{star.logs[,length(unique(file))]} term placentas of
  \Sexpr{star.logs[,length(unique(species))]} species of mammals 
  also identified genes which are expressed in all placentas,
  including genes with immune-modulating properties (\textit{ANXA2},
  \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}) and
  cell-cell interactions (\textit{LAMC1}, \textit{LUM}, and
  \textit{LGALS1}).

  \textbf{Discussion}

  Still, using the strictest filtering strategy
  we find a core set of only
  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
  genes is essential for the proper functioning of the placenta in all
  species analyzed.

@@

\section{Introduction}

@@

@@

\subsection{Pre-eclampsia \human Lineage Specific Genes}

@@
